\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-22 forwards-and-backwards.tex
.

Putting these together,
the level $u$ of an individual at $x$ in $\lp$ evolves according to:
\begin{equation} \label{eqn:dot_u}
    \dot u
    =
    - \frac{\theta}{\lambda} u \left(\lambda - u\right)
    \gamma(x, \eta) \int_{\IR^d} r(y, \eta) q(x, dy) 
    +
    \theta \mu(x,\eta) u .
\end{equation}
The factor of $\theta$ appears here because we have rescaled time.
We shall write 
$$
    b(x, \eta)
    :=
    \theta\left(
    \gamma(x,\eta) \int_{\IR^d} r(y, \eta) q(x, dy)
    -
    \mu(x,\eta)
    \right) ,
$$
which captures the local net difference between reproduction and death,
i.e., the deviation from criticality of the branching mechanism.
Recall from equation \eqref{eqn:mu_defn} that
$F(x,\eta) = \theta(r(x,\eta)\gamma(x,\eta) - \mu(x,\eta))$,
and so we will have that
\begin{align} \label{eqn:b_limit}
b(x, \eta)
&=
    \theta \gamma(x, \eta) \int_{\IR^d} \left( r(y, \eta) - r(x, \eta) \right) q(x, dy)
    + F(x, \eta) \\
&\to
    \gamma(x, \eta) \DG r(x, \eta) + F(x, \eta) \qquad \text{as } \theta \to \infty .
\end{align}

@@

@@

The main interest in using a lookdown construction for our 
population processes is that it allows us to retain information 
about relatedness between individuals as we pass to the infinite
density limit. In order to exploit this, we generalise the construction of 
stochastic equations presented in \S \ref{sec: Evolution of an Individual Particle} to the infinite density limit.\\
In the scaling limit $A^{\text{nl}}_{\epsilon}$, the spatial movement of particles can be characterised with the map $$h \mapsto \gamma\big(x,(\rho_{\epsilon}*\Xi)(x)\big)r\big(x,(\rho_{\epsilon}*\Xi)(x)\big)\mathcal{B}^{m}_{\epsilon}(\Xi)h(x) ,$$
and as individual particles travel as a random walk in the pre-limit model, we should expect a diffusion limit for individual trajectories.
\todo[inline]{From Terence: we should try explain in detail why this is the generator of an individual line of descent forward-in-time. The main question here is whether we should have a factor of $r$ in the above notation. For the proof to work in \S \ref{sec: motion of ancestral lineages} we will need a factor of $r$ but I have been finding it slightly unintuitive particularly since the effect of $r$ has been accounted into $\mathcal{B}^{\mathfrak{m}}$ through the limit.}
To simply notation, we let the above map be the generator of a process 
satisfying 
\[X(t)=X(0)+\int_0^t\sigma (X(s),\Xi (s))dW(s)+\int\beta (X(s),\Xi 
(s))ds.\]
\begin{remark}
In particular recall that as in Remark \ref{Remark: Product Rule on Laplacian}, if $\mathcal{B}$ is the Laplacian operator, then the generator above is merely 
$$\gamma\big(x,(\rho_{\epsilon}*\Xi)(x)\big)\left(\Delta h(x) +2 \nabla h(x) \cdot \nabla \log r( \cdot, (\rho_{\epsilon}*\Xi)(\cdot))(x)\right),$$
and applying Ito's formula we have 
\begin{equation}\label{eq: Individual lineage generator description}
\begin{aligned}
\beta(X(s),\Xi(s))=& 2\gamma\big(X(s),(\rho_{\epsilon}*\Xi)(X(s))\big) \nabla \log r (\cdot, (\rho_{\epsilon}*\Xi)(\cdot))(X(s))\\
\sigma(X(s),\Xi(s))=& \sqrt{2\gamma\big(X(s),(\rho_{\epsilon}*\Xi)(X(s))\big)}.
\end{aligned}
\end{equation}
\end{remark}
Furthermore, recall that in our scaling limit, 
\begin{equation}
\begin{aligned}
\lambda^{-1} \theta \gamma^{\mathfrak{m}}_{\theta}(X_{\tau}(s), \eta(s)) &\to \ell \gamma(X_{\tau}(s),\rho_\epsilon*\Xi(X_{\tau}(s)),\\
b_{\theta}(X_{\tau}(s),\eta(s)) &\to  F(X_{\tau}(s), (\rho_{\epsilon}*\eta)(X_{\tau}(s)))\\
&+\gamma(X_{\tau}(s), (\rho_{\epsilon}*\Xi)(X_{\tau}(s)))\mathcal{B}r(\cdot,(\rho_{\epsilon}*\eta)(\cdot))(X_{\tau}(s)).
\end{aligned}    
\end{equation}

@@

First, the contribution from births is
\begin{align*}
\begin{split}
f(\lp)
&\mapsto
    f(\lp)
    \sum_{(x, u) \in \lp}
    2 \gamma(x, \eta)
    \bigg\{
        \frac{1}{2 \lambda}
        \int_u^\lambda
        g(x, u_1) du_1
        \frac{
            \theta \int_{\IR^d} (g(y, u) - g(x, u)) r(y, \eta) q(x, dy)
        }{
            g(x, u)
        }
    \\ & \qquad \qquad \qquad {}
        + \frac{\theta}{\lambda}
        \int_u^\lambda \int_{\IR^d}
        \left( \frac{g(y, u_1) + g(x, u_1)}{2} - 1 \right)
        r(y, \eta) q(x, dy)
    \bigg\}
    .
    \end{split}
\end{align*}
We have that
\begin{align*}
    \theta \int_{\IR^d} (g(y, u) - g(x, u)) r(y, \eta) q(x, dy)
    \to
    \DG(gr)(x) \qquad \text{as }\theta \to \infty,
\end{align*}
i.e., to $\DG\left(g(\cdot, u) r(\cdot, \smooth{r}\eta(\cdot))\right)(x)$.
Furthermore, by our assumption on $g$,
$\int_u^\lambda g(x, u_1) du_1 / \lambda \to 1$ for all $x$ and $u$,
so the first term here converges to
\begin{align*}
    \gamma(x, \eta)
        \frac{
            \DG\left( g(\cdot, u) r(\cdot, \smooth{r}\eta(\cdot))\right)(x)
        }{
            g(x, u)
        } .
\end{align*}
This corresponds to each particle moving according to the motion induced by
$g \mapsto \gamma \DG(gr)$.
\comment{TODO: explain the $r$-tilted $\DG$ motion somewhere.}
As for the second term,
since $\int g(y, u_1) q(x, dy) \to g(x, u_1)$
and $\theta/\lambda \to \delta$,
it converges to
\begin{align*}
    2 \delta
    \gamma(x, \eta)
    r(x, \eta) 
    \int_u^\infty
    \left( g(x, u_1) - 1 \right)
    du_1
\end{align*}
This corresponds to births at higher levels at rate $2 \delta \gamma r$.

>>>>

Putting these together,
the level $u$ of an individual at $x$ in $\lp$ evolves according to:
\begin{equation} \label{eqn:dot_u}
    \dot u
    =
    - \frac{\theta}{\lambda} u \left(\lambda - u\right)
    \gamma(x, \eta) \int_{\IR^d} r(y, \eta) q(x, dy) 
    +
    \theta \mu(x,\eta) u .
\end{equation}
The factor of $\theta$ appears here because we have rescaled time.
We shall write 
$$
    b(x, \eta)
    :=
    \theta\left(
    \gamma(x,\eta) \int_{\IR^d} r(y, \eta) q(x, dy)
    -
    \mu(x,\eta)
    \right) ,
$$
which captures the local net difference between reproduction and death,
i.e., the deviation from criticality of the branching mechanism.
Recall from equation \eqref{eqn:mu_defn} that
$F(x,\eta) = \theta(r(x,\eta)\gamma(x,\eta) - \mu(x,\eta))$,
and so we will have that
\begin{align} \label{eqn:b_limit}
\begin{split}
b(x, \eta)
&=
    \theta \gamma(x, \eta) \int_{\IR^d} \left( r(y, \eta) - r(x, \eta) \right) q(x, dy)
    + F(x, \eta) \\
&\to
    \gamma(x, \eta) \DG r(x, \eta) + F(x, \eta) \qquad \text{as } \theta \to \infty .
\end{split}
\end{align}

@@

The only thing that remains is to describe how the levels change.
The limit of~\eqref{eqn:line_points} follows from
equation \eqref{eqn:b_limit} and the fact that
$\int_{\IR^d} r(z, \eta) q(x, dz) \to r(x, \eta)$.

To write out the joint motion of spatial coordinate and levels,
it is useful to work out the differential operator above in more detail.
Recall that $\DG g(x) = \sum_i \meanq_i \partial_i g(x) + \sum_{ij} \covq_{ij} \partial_{ij} g(x)$,
and for the moment write $r(x)$ for $r(\cdot, \eta(\cdot))(x)$,
so that
\begin{align*}
\DG(rg)(x) - g(x) \DG(r)(x)
    &=
    r(x) \sum_i \meanq_i \partial_i g(x)
    + 2 \sum_{ij} \meanq_i \partial_i r(x) \covq_{ij} \partial_j g(x)
    + r(x) \sum_{ij} \covq_{ij} \partial_{ij} g(x) \\
    &=
    r(x) \left\{
        \left(
        \meanq
        + 2 \covq \grad \log r(x)
        \right)
        \cdot
        \grad g(x)
        +
        \sum_{ij} \covq_{ij} \partial_{ij} g(x)
    \right\} .
\end{align*}

Now, let $\beta(x, \eta) = \meanq + 2 \Sigma \grad \log r(x, \eta)$,
and associate with each label an independent $d$-dimensional Brownian motion $W_a$,
and let $\covq = K K^T$.
Then, we can summarize the results above by saying that
in the infinite density limit,
the spatial locations and level of each line of descent $a$
behave for $\tau_a \le t < \sigma_a$ as
\begin{equation*}
\begin{split}
X_a(t)
    &=
    X_a(\tau_a)
    +
    + \int_{\tau_a}^{t}
    \bigg(
        \beta(x, \eta(s)) ds
        +
        K dW_a(s)
    \bigg)
    ds .
\end{split}
\begin{split}
U_a(t)
    &=
    U_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \delta \gamma(X_a(s),\eta(s))
        r(x,\eta(s)) U_a(s)^2
\\ &\qquad \qquad \qquad {}   
        -
        \left\{
            \gamma(X_a(s),\eta(s)) \DG r(X_a(s),\eta(s))
            + F(X_a(s), \eta(s))
        \right\}
        U_a(s)
    \bigg)
    ds .
\end{split}
\end{equation*}
\comment{TODO: add description of Poisson random measure of birth times/levels as below.}

@@

The main interest in using a lookdown construction for our 
population processes is that it allows us to retain information 
about relatedness between individuals as we pass to the infinite
density limit. In order to exploit this, we generalise the construction of 
stochastic equations presented in \S \ref{sec: Evolution of an Individual Particle} to the infinite density limit.\\
In the scaling limit $A^{\text{nl}}_{\epsilon}$, the spatial movement of particles can be characterised with the map $$h \mapsto \gamma\big(x,(\rho_{\epsilon}*\Xi)(x)\big)r\big(x,(\rho_{\epsilon}*\Xi)(x)\big)\mathcal{B}^{m}_{\epsilon}(\Xi)h(x) ,$$
and as individual particles travel as a random walk in the pre-limit model, we should expect a diffusion limit for individual trajectories.
\todo[inline]{From Terence: we should try explain in detail why this is the generator of an individual line of descent forward-in-time. The main question here is whether we should have a factor of $r$ in the above notation. For the proof to work in \S \ref{sec: motion of ancestral lineages} we will need a factor of $r$ but I have been finding it slightly unintuitive particularly since the effect of $r$ has been accounted into $\mathcal{B}^{\mathfrak{m}}$ through the limit.}
To simply notation, we let the above map be the generator of a process 
satisfying 
\[X(t)=X(0)+\int_0^t\sigma (X(s),\Xi (s))dW(s)+\int\beta (X(s),\Xi 
(s))ds.\]
\begin{remark}
In particular recall that as in Remark \ref{Remark: Product Rule on Laplacian}, if $\mathcal{B}$ is the Laplacian operator, then the generator above is merely 
$$\gamma\big(x,(\rho_{\epsilon}*\Xi)(x)\big)\left(\Delta h(x) +2 \nabla h(x) \cdot \nabla \log r( \cdot, (\rho_{\epsilon}*\Xi)(\cdot))(x)\right),$$
and applying Ito's formula we have 
\begin{equation}\label{eq: Individual lineage generator description}
\begin{aligned}
\beta(X(s),\Xi(s))=& 2\gamma\big(X(s),(\rho_{\epsilon}*\Xi)(X(s))\big) \nabla \log r (\cdot, (\rho_{\epsilon}*\Xi)(\cdot))(X(s))\\
\sigma(X(s),\Xi(s))=& \sqrt{2\gamma\big(X(s),(\rho_{\epsilon}*\Xi)(X(s))\big)}.
\end{aligned}
\end{equation}
\end{remark}

Furthermore, recall that in our scaling limit, 
\begin{equation}
\begin{aligned}
\lambda^{-1} \theta \gamma^{\mathfrak{m}}_{\theta}(X_{\tau}(s), \eta(s)) &\to \ell \gamma(X_{\tau}(s),\rho_\epsilon*\Xi(X_{\tau}(s)),\\
b_{\theta}(X_{\tau}(s),\eta(s)) &\to  F(X_{\tau}(s), (\rho_{\epsilon}*\eta)(X_{\tau}(s)))\\
&+\gamma(X_{\tau}(s), (\rho_{\epsilon}*\Xi)(X_{\tau}(s)))\mathcal{B}r(\cdot,(\rho_{\epsilon}*\eta)(\cdot))(X_{\tau}(s)).
\end{aligned}    
\end{equation}

@@

First, the contribution from births is
\begin{align*}
\begin{split}
f(\lp)
&\mapsto
    f(\lp)
    \sum_{(x, u) \in \lp}
    2 \gamma(x, \eta)
    \bigg\{
        \frac{1}{2 \lambda}
        \int_u^\lambda
        g(x, u_1) du_1
        \frac{
            \theta \int_{\IR^d} (g(y, u) - g(x, u)) r(y, \eta) q(x, dy)
        }{
            g(x, u)
        }
    \\ & \qquad \qquad \qquad {}
        + \frac{\theta}{\lambda}
        \int_u^\lambda \int_{\IR^d}
        \left( \frac{g(y, u_1) + g(x, u_1)}{2} - 1 \right)
        r(y, \eta) q(x, dy)
    \bigg\}
    .
    \end{split}
\end{align*}
We have that
\begin{align*}
    &\theta \int_{\IR^d} (g(y, u) - g(x, u)) r(y, \eta) q(x, dy) \\
    &\qquad =
    \theta \int_{\IR^d} (r(y, \eta) g(y, u) - r(x, \eta) g(x, u)) q(x, dy)
    - \theta \int_{\IR^d} g(x, u) (r(y, \eta) - r(x, \eta)) q(x, dy) \\
    &\qquad \to
    \DG(gr)(x) - g(x) \DG r(x) \qquad \text{as }\theta \to \infty.
\end{align*}
i.e., to
$\DG\left(g(\cdot, u) r(\cdot, \smooth{r}\eta(\cdot))\right)(x)
- g(\cdot, u) \DG\left(r(\cdot, \smooth{r}\eta(\cdot))\right)(x)$.
Furthermore, by our assumption on $g$,
$\int_u^\lambda g(x, u_1) du_1 / \lambda \to 1$ for all $x$ and $u$,
so the first term here converges to
\begin{align*}
    \gamma(x, \eta)
        \frac{
            \DG\left( g(\cdot, u) r(\cdot, \smooth{r}\eta(\cdot))\right)(x)
            -
            g(\cdot, x) \DG\left( r(\cdot, \smooth{r}\eta(\cdot))\right)(x)
        }{
            g(x, u)
        } .
\end{align*}
This corresponds to each particle moving according to the motion induced by
$g \mapsto \gamma \DG(gr) - g \DG r$.
\comment{TODO: explain the $r$-tilted $\DG$ motion somewhere earlier, introducing notation for it.}
As for the second term,
since $\int g(y, u_1) q(x, dy) \to g(x, u_1)$
and $\theta/\lambda \to \delta$,
it converges to
\begin{align*}
    2 \delta
    \gamma(x, \eta)
    r(x, \eta) 
    \int_u^\infty
    \left( g(x, u_1) - 1 \right)
    du_1
\end{align*}
This corresponds to births at higher levels at rate $2 \delta \gamma r$.
